---
title: "Data 205 EDA"
format: html
editor: visual
---

```{r}
library(tidyverse)
setwd("/Users/blossomanyanwu/Documents/Data 205 Project")
```

```{r}
countyexercise<-read_csv("exercisecounty.csv")
internetaccess<-read_csv("internet.csv")
housingage<-read_csv("1980housing.csv")
usainternet<-read_csv("usainternet.csv")
usahousing<-read_csv("countyhousing.csv")
```

```{r}
# Merge USA internet data and USA Housing Data
merged_data <- usainternet|>
  left_join(usahousing, by = c(StateFIPS = "StateFIPS", State = "State", 
                         CountyFIPS = "CountyFIPS",`Data Comment` = "Data Comment", '...9' = "...9", County = "County"))
# Remove empty columns 
merged_data <- merged_data %>%
  select(-c('...9', 'Data Comment'))
```

# Subset DMV Data 
```{r}
# Subset data for all target states
target_states <- c("District of Columbia", "Maryland", "Virginia")
dmvdata <- merged_data[merged_data$State %in% target_states, ]
```


```{r}
ggplot(countyexercise, aes(x = County, y = Value)) +
  geom_bar(stat = "identity", fill = "skyblue") +  # Adjust bar color
  labs(title = "Physical Inactivity Rates by County Type",
       x = "County Type",
       y = "% Physically Inactive", 
       caption = "Source: CDC Public Health Tracking Network (accessed April 2, 2024)") +
  theme_classic()  # One can adjust theme as desired
```


# Bar Plot of Region Internet Access
```{r}
ggplot(dmvdata, aes(x = County, y = Number)) +
  geom_bar(stat = "identity", fill = "skyblue") +  # Adjust bar color
  labs(title = "Internet",
       x = "County",
       y = "Number of Housing built prior to 1980", 
       caption = "Source: CDC Public Health Tracking Network (accessed April 2, 2024)") +
  theme_classic()  # One can adjust theme as desired
```
# Histogram of Households without Internet
```{r}

ggplot(dmvdata, aes(x = Number)) +
  geom_histogram(bins = 30, color = "lightblue") +  # Adjust 'bins' for desired number of bars
  labs(title = "Histogram of Income", x = "Income", y = "Frequency") +
  theme_classic()
```



# Merge Housing Data and Internet Access data set (Operates as Community Design Data set)
```{r}

combined <- housingage|>
  left_join(internetaccess, by = c(StateFIPS = "StateFIPS", State = "State", 
                         CensusTract = "CensusTract", `Census Tract` = "Census Tract",`Data Comment` = "Data Comment"))
```


# Scatter Plot: Housing Built before 1980 and Internet Access
```{r}
### Does housing age impact technical infrastructure and access to tech?
ggplot(combined, aes(x = Value.x, y = Value.y)) +
 geom_point(size = 3) +  # Adjust point size and add color mapping (optional)
  labs(title = "Housing Age and Internet Access",
       x = "Percent of Housing Built Prior to 1980",
       y = "Number of Households Without Intern Access") +
  theme_classic()
```
```{r}
usainternet<-read_csv("usainternet.csv")
usahousing<-read_csv("countyhousing.csv")
```


# EDA Continued
```{r}

options(scipen = 999)
ggplot(merged_data, aes(x = Number, y = Value.x)) +
 geom_point(size = .5) +  # Adjust point size and add color mapping (optional)
  labs(title = "Housing Age and Internet Access",
       x = "Number of Housing Units Built before 1950",
       y = "Number of Households Without Internet Access") +
  theme_classic()
```

# Linear Model of All USA Counties
```{r}

model <- lm(Number ~ Value.x, data = merged_data)  # y ~ x represents dependent variable ~ independent
summary(model)
```
# Scatter Plot of Singular Region (DMV)
```{r}

# Create the scatter plot
ggplot(dmvdata, aes(x = Number, y = Value.x)) +
  geom_point(color = "blue", alpha = 0.5) +  # Adjust color and transparency as desired
  labs(title = paste0("Scatter Plot for DC, MD, VA:"),
       x = "Number of Households without Internet Access", y = "Number of Housing Units built before 1980") +  # Customize axis labels
  theme_classic()  # Adjust theme (optional)
```
# Linear Model of One Region (DMV)
```{r}
model2 <- lm(Number ~ Value.x, data = dmvdata)  # y ~ x represents dependent variable ~ independent
summary(model2)
```
# Maryland Specific Analysis 
```{r}
target_county <- "Maryland"

# Subset data for the target county
state_data <- merged_data[merged_data$State == target_county, ]

# Create a state specific scatter plot 
ggplot(state_data, aes(x = Number, y = Value.x)) +
  geom_point(color = "blue", alpha = 0.5) +  # Adjust color and transparency as desired
  labs(title = paste0("Scatter Plot for Maryland:"),
       x = "Number of Housing Units built before 1980", y = "Number of Households without Internet Access") +
  theme_minimal()
```



# Notes: Models and visualization have established a significant positive linear correlation between number of housing units built prior to 1980 and number of housholds without intern access. Now one can move towards exploring how this situation impact people living nearby by. The chosen health focuses are Access to Food, Mental Health, and Obesity











